Mechanization of Separation in Generic Extensions

### 2019-04-01 ITP/Separation_In_MG.tex
Cambios de acuerdo al review 4 de LICS

\acknowledgements{We would like to thanks to anonymous referees of a
  previous version of this paper that helped us to enhance the
  presentation.}%optional

@@

\newpage
\onecolumn
\appendix[A short overview of our development]
\input{appendix}
% that's all folks
\end{document}

>>>>

\acknowledgements{We would like to thanks to anonymous referees of a
  previous version of this paper that helped us to enhance the
  presentation. We are grateful to Alexandre Miquel for his comments
  on forcing and Boolean-valued models.}%optional

@@

\newpage
\onecolumn
\appendix[A short overview of our development]
\input{appendix}

\end{document}

### 2019-04-01 ITP/conclusion.tex
Cambios de acuerdo al review 4 de LICS

A secondary, more prosaic, outcome of this project is to precisely
assess which assumptions on the ground model $M$ are needed to develop
the forcing machinery. The obvious are transitivity and $M$ being
countable (but keep in mind Lemma~\ref{lem:wf-model-implies-ctm}); the
first because many absoluteness results follows from this, the latter
for the existence of generic filters. A
more anecdotal one is that to show that an instance of Separation
with at most two parameters holds in $M[G]$, one needs to assume a
particular  six-parameter
instance in $M$ (four extra parameters can be directly blamed on
$\forceisa$).  The purpose of identifying those assumptions is to assemble
in a locale the specific (instances of) axioms that should satisfy the
ground model in order to perform forcing constructions; this list will
likely include all the instances of Separation and Replacement that
are needed to satisfy the requirements of the locales in the
\isatt{ZF-Constructible} session.

@@

>>>>

A secondary, more prosaic, outcome of this project is to precisely
assess which assumptions on the ground model $M$ are needed to develop
the forcing machinery. The obvious are transitivity and $M$ being
countable (but keep in mind Lemma~\ref{lem:wf-model-implies-ctm}); the
first because many absoluteness results follows from this, the latter
for the existence of generic filters. % A
% more anecdotal one is that to show that an instance of Separation
% with at most two parameters holds in $M[G]$, one needs to assume a
% particular  six-parameter
% instance in $M$ (four extra parameters can be directly blamed on
% $\forceisa$).
The purpose of identifying those assumptions is to assemble
in a locale the specific (instances of) axioms that should satisfy the
ground model in order to perform forcing constructions; this list will
likely include all the instances of Separation and Replacement that
are needed to satisfy the requirements of the locales in the
\isatt{ZF-Constructible} session.

@@

Recently Alexandre Miquel suggested us to look into Boolean-valued
models as an alternative method for extending models of ZF. This is a
very interesting option to consider, since that would work for any
model (and not just transitivive ones); some of the others issues (the
most salient being the use of internalized formulas) are consequences
of working in Isabelle/ZF and would be inevitable.

### 2019-04-01 ITP/easy-axioms.tex
Cambios de acuerdo al review 4 de LICS

In our first presentation of this project \cite{2018arXiv180705174G},
we proved that $M[G]$ satisfies Pairing; now we have redone this proof
in Isar. It is straightforward to show that the generic extension
$M[G]$ satisfies extensionality and foundation. Showing that it is
closed under Union depends on $G$ being a filter. Infinity is also
easy, but it depends in one further assumption.

>>>>

In this section we briefly comment the proofs showing that the generic
extension $M[G]$ satisfies extensionality and foundation. Showing that
it is closed under Union depends on $G$ being a filter. Infinity is
also easy, but it depends in one further assumption.

### 2019-04-01 ITP/intro.tex
Cambios de acuerdo al review 4 de LICS

It might be a little surprising the lack of formalizations of forcing
and generic extensions. As far as we know, the development of
Quirin and Tabareau \cite{JFR6232} in homotopy type theory for constructing generic
extensions in a sheaf-theoretic setting is the unique mechanization of
forcing. This contrast with the fruitful use of forcing techniques to
extend the Curry-Howard isomorphism to classical axioms
\cite{Miquel:2011:FPT:2058525.2059614,lmcs:1070}. Moreover, the
combination of forcing with intuitionistic type theory
\cite{Coquand:2009:FTT:1807662.1807665,coquand2010note} gives rise
both to positive results (an algorithm to obtain witnesses of the
continuity of definable functionals \cite{coquand2012computational})
and also negative (the independence of Markov's principle
\cite{lmcs:3859}). In the same strand of forcing from the point of
view of proof theory \cite{avigad_2004} are the conservative
extensions of CoC with forcing conditions
\cite{jaber:hal-01319066,Jaber:2012:ETT:2358958.2359524}.

@@

In pursuing the proof of Separation on generic extensions we
extended Paulson's library with:
\begin{enumerate}
\item renaming of variables for \emph{internalized} formulas, which
  with little effort can be extended 
  to substitutions;
\item an improvement on definitions by recursion on well-founded
  relations; 
\item enhancements in the hierarchy of locales; and
\item a variant of the  principle of dependent choices and a version
  of Rasiowa-Sikorski's theorem, which 
  ensures the existence of generic filters for countable and transitive
  models of $\ZF$;
\end{enumerate}
the last item was already communicated in the
  first report \cite{2018arXiv180705174G}.

>>>>

It might be a little surprising the lack of formalizations of forcing
and generic extensions. As far as we know, the development of
Quirin and Tabareau \cite{JFR6232} in homotopy type theory for constructing generic
extensions in a sheaf-theoretic setting is the unique mechanization of
forcing. This contrast with the fruitful use of forcing techniques to
extend the Curry-Howard isomorphism to classical axioms in order to
get programs for proofs using those axioms
\cite{Miquel:2011:FPT:2058525.2059614,lmcs:1070}. Moreover, the
combination of forcing with intuitionistic type theory
\cite{Coquand:2009:FTT:1807662.1807665,coquand2010note} gives rise
both to positive results (an algorithm to obtain witnesses of the
continuity of definable functionals \cite{coquand2012computational})
and also negative (the independence of Markov's principle
\cite{lmcs:3859}). In the same strand of forcing from the point of
view of proof theory \cite{avigad_2004} are the conservative
extensions of CoC with forcing conditions
\cite{jaber:hal-01319066,Jaber:2012:ETT:2358958.2359524}.

@@

In pursuing the proof of Separation on generic extensions we
extended Paulson's library with:
\begin{enumerate}
\item renaming of variables for \emph{internalized} formulas, which
  with little effort can be extended 
  to substitutions;
\item an improvement on definitions by recursion on well-founded
  relations; 
\item enhancements in the hierarchy of locales; and
\item a variant of the  principle of dependent choices and a version
  of Rasiowa-Sikorski's theorem, which 
  ensures the existence of generic filters for countable and transitive
  models of $\ZF$;
\end{enumerate}
the last item was already communicated in the first report
\cite{2018arXiv180705174G} of our project (where we also proved that
$M[G]$ satisfies Pairing; now we have redone this proof in Isar).

### 2019-04-01 ITP/isabelle.tex
Terminando con todas las observaciones del review 1(?)
Cambios de acuerdo al review 4 de LICS

We work in the object theory \emph{Isabelle/ZF}. Two types are defined
in this theory: \tyo, the object-Booleans, and \tyi,
sets. 
%% It must be insisted that the types are defined axiomatically, not
%% recursively. That is, although there are constants and functions that
%% generate elements of both types, neither of them is 
%% \emph{initial}, in the sense that they are not the least types
%% obtained by combining the given constants and operations. 
It must be observed that predicates (functions with arguments of
type \tyi{} with values in \tyo) do not correspond to first-order formulas;
in particular, those are not recursively constructed.
This will
have concrete consequences in our strategy to approach the
development. From the beginning, we had to resort to
\emph{internalized} formulas \cite[Sect.~6]{paulson_2003}, i.e.\ elements of type $\tyi$ that
encode first-order formulas with a binary relation symbol, and the
satisfaction predicate \isatt{sats\,::\,"i\isasymRightarrow i\isasymRightarrow i\isasymRightarrow o"}  between a set
model with an environment and an internalized formula (where the
relation symbol is interpreted as membership). The set 
\isatt{formula::"\tyi"}
 of internalized
formulas is inductively defined and hence it is possible to perform
inductive arguments on them, such as the definition of the renaming map. In this sense, the object-logic level
is further divided into \emph{internal} and \emph{external}
sublevels.

>>>>

We work in the object theory \emph{Isabelle/ZF}. Two types are defined
in this theory: \tyo, the object-Booleans, and \tyi,
sets. 
%% It must be insisted that the types are defined axiomatically, not
%% recursively. That is, although there are constants and functions that
%% generate elements of both types, neither of them is 
%% \emph{initial}, in the sense that they are not the least types
%% obtained by combining the given constants and operations. 
It must be observed that predicates over \tyi{} do not correspond to
first-order formulas; in particular, those are not recursively
constructed and there is no induction principle associated with them.
This will have concrete consequences in our strategy to approach the
development. From the beginning, we had to resort to
\emph{internalized} formulas \cite[Sect.~6]{paulson_2003}, i.e.\
elements of type $\tyi$ that encode first-order formulas with a binary
relation symbol, and the satisfaction predicate
\isatt{sats\,::\,"i\isasymRightarrow i\isasymRightarrow
  i\isasymRightarrow o"} between a set model with an environment and
an internalized formula (where the relation symbol is interpreted as
membership). The set \isatt{formula::"\tyi"} of internalized formulas
is inductively defined and hence it is possible to perform inductive
arguments on them, such as the definition of the renaming map. In this
sense, the object-logic level is further divided into \emph{internal}
and \emph{external} sublevels. We remark that Isabelle/ZF lacks any
mechanism to define inductively defined types at the same level of
\tyi{} or \tyo.
